Tidal streams in triaxial systems

### 2015-03-22 ms.tex
added monte carlo blip
explain the projection of distribution function better

NAFF recovers the fundamental frequencies for an orbit faster (with a fewer number of terms) when the coordinates used are ``close'' to the angle variables \cite[PL96][]{papaphilippou96}. PL96 show that a good choice of coordinates for tube orbits are the Poincar\'e symplectic polar coordinates, a set of canonical coordinates similar to cylindrical coordinates. When computing the frequencies for tube orbits, we first rotated the orbits until the circulation is about the $z$-axis, transform to Poincar\'e polar coordinates, then use NAFF to measure the fundamental frequencies. We could equivalently use the Cartesian time series, but the convergence of terms is slower. We have tested that our implementation of NAFF returns the same frequencies in either case for a set of tube orbits. For box orbits, the motion is close to separable in each Cartesian component and we therefore use Cartesian coordinates for estimating frequencies for these orbits.

@@

The Lyapunov time and frequency diffusion rate measure the timescales over which chaos is relevant \emph{local} to a given orbit. That is, they are measures of how infinitesimal deviations will diverge, on average, from some parent orbit of interest. Tidal debris is disrupted from progenitor systems with finite spreads in orbital properties (e.g., energy). For a disrupting, globular-cluster-scale progenitor, the typical energy dispersion of the debris is 0.1--1\% of the progenitor orbital energy, but for a dwarf-galaxy-scale progenitor, the dispersion can be 5--20\%. This raises an important consideration: do finite deviations diverge faster than predicted from local measures of chaos? Does the Lyapunov time or frequency rate predict the timescale over which a finite phase-space volume stays ``coherent''?

It is computationally intractable to run full N-body simulations of stellar system disruption for large grids of orbital initial conditions and thus we take a simplified approach for analyzing these questions. We instead consider small ensembles of particles meant to represent debris disrupted from a single tidal disruption event. For a given set of orbital initial conditions --- the parent orbit --- we find the position of the nearest pericenter, initialize a small ensemble of test particle orbits around this position, then follow the orbits of all of these test particles for some evolution time, $T_{\rm ev}$. The physical scale of the ensemble is set by the tidal radius in position and velocity scale in velocity and are therefore set by the mass scale of the progenitor \citep[e.g.,][]{johnston98, apw14}. If $(\bs{x}_0,\bs{v}_0)$ are the parent orbit initial conditions at pericenter, then $(\delta\bs{x}_i,\delta\bs{v}_i)$ is the deviation vector of the $i$th particle and the magnitude of the offsets are assumed to be Normally distributed away from the parent orbit:
\begin{align}
	\delta\bs{x}_i &\sim \mathcal{N}(0, r_{\rm tide})\\
	\delta\bs{v}_i &\sim \mathcal{N}(0, \sigma_v)\\
	r_{\rm tide} &= \|\bs{x}_0\| \left(\frac{m}{3M_{\rm enc}(\|\bs{x}_0\|)}\right)^{1/3} \\
	\sigma_v &= \|\bs{v}_0\|\left(\frac{m}{3M_{\rm enc}(\|\bs{x}_0\|)}\right)^{1/3} 
\end{align}
where $M_{\rm enc}(r)$ is the mass enclosed of the host potential within radius $r$, $m$ is the mass scale of the debris, and $\|\cdot \|$ is the Euclidean norm. For this work, we take $m=2\times10^4~\msun$ to represent globular-cluster-like progenitors, and use the spherically-averaged enclosed mass to estimate the above scales.

As the particle ensembles evolve in time, we follow a method similar to that of \citet[][hereafter MV96]{merritt96} and quantify the degree of mixing of the ensemble by computing the ``distance'' between the configuration space density of the ensemble at a given time, $\rho(\bs{x},t)$, with that of a distribution that is uniform over the energy hypersurface, $\rho_{E_0}(\bs{x})$, where $E_0$ is the energy of the parent orbit and thus hypersurface of interest. Our method differs in a few ways from that of MV96. First, MV96 use isoenergy ensembles of particles for following the density because this is the only way to guarantee that, if chaotic, nearly all orbits in the ensemble will lie on chaotic orbits. We instead use ensembles with small, finite spreads in energy, but we still expect that most ensemble particle orbits around a parent chaotic orbit will reach a fully mixed state. The final distribution, in addition to filling the energy hypersurface, will also have some thickness around this surface. Around a typical regular orbit, each individual orbit will fill its torus ergodically --- a surface with lower dimension than the energy hypersurface --- and thus, a small enough ensemble will never reach the fully mixed state as defined above because it will remain confined into a subspace of lower dimensionality. In practice, only strongly chaotic systems will get close to $\rho_{E_0}(\bs{x})$; we expect the rate to depend on how stochastic the parent orbit is (e.g., as measured by the Lyapunov time or frequency diffusion rate) and also on how ``trapped'' the chaotic region is (a small chaotic region wedged between stable tori will take longer to appear chaotic, even if internally the frequency diffusion rate is large). However the difference in configuration space density between an ensemble undergoing chaotic mixing versus an ensemble around a regular orbit may be detectable over a much shorter timescale. The fully-mixed, normalized, configuration-space density that fills a given energy hypersurface, $E_0$, is 
\begin{align}
	\rho_{E_0}(\bs{x}) &= \int \dd^3v \, \rho_0 \, \delta(H - E_0)\\
	&= \int \dd^3v \, \rho_0 \, \delta\left(|\bs{v}|^2/2 + \Phi(\bs{x}) - E_0\right)\\
	&= 4\pi \int \dd v \, v^2 \, \rho_0 \, \delta\left(v^2/2 + \Phi(\bs{x}) - E_0\right)\\
	&= A \sqrt{E_0 - \Phi(\bs{x})}\label{eq:mixeddensity}
\end{align}
where $A$ is a constant determined such that $\int \dd^3\bs{x} \, \rho_{E_0}(\bs{x}) = 1$, and $\delta(\cdot)$ is the Dirac delta function.

@@

>>>>

NAFF recovers the fundamental frequencies for an orbit faster (with a fewer number of terms) when the coordinates used are ``close'' to the angle variables \cite[PL96][]{papaphilippou96}. PL96 show that a good choice of coordinates for tube orbits are the Poincar\'e symplectic polar coordinates, a set of canonical coordinates similar to cylindrical coordinates. When computing the frequencies for tube orbits, we first rotate the orbits until the circulation is about the $z$-axis, transform to Poincar\'e polar coordinates, then use NAFF to measure the fundamental frequencies. We could equivalently use the Cartesian time series, but the convergence of terms is slower (the amplitudes of successive terms decrease slower for Cartesian coordinates). We have tested that our implementation of NAFF returns the same fundamental frequencies in either case for a set of tube orbits. For box orbits, the motion is close to separable in each Cartesian component and we therefore use Cartesian coordinates for estimating frequencies for these orbits.

@@

The Lyapunov time and frequency diffusion rate measure the timescales over which chaos is relevant \emph{local} to a given orbit --- that is, these quantities are measures of how infinitesimal deviations will diverge, on average, from some parent orbit of interest. Tidal debris is disrupted from progenitor systems with finite spreads in orbital properties (e.g., energy). For a disrupting, globular-cluster-scale progenitor, the typical energy dispersion of the debris is 0.1--1\% of the progenitor orbital energy, but for a dwarf-galaxy-scale progenitor, the dispersion can be 5--20\%. This raises an important consideration: do finite deviations diverge faster than predicted from local measures of chaos? Does the Lyapunov time or frequency rate predict the timescale over which a finite phase-space volume stays ``coherent''?

It is computationally intractable to run full N-body simulations of stellar system disruption for large grids of orbital initial conditions and thus we take a simplified approach for analyzing these questions. We instead consider small ensembles of particles meant to represent debris disrupted from a single tidal disruption event. For a given set of orbital initial conditions --- the ``parent'' orbit --- we find the position of the nearest pericenter, initialize a small ensemble of test particle orbits around this position, then follow the orbits of all of these test particles for an integration time equal to 256 times the longest fundamental orbital period. The physical scale of the ensemble is set by the tidal radius in position and velocity scale in velocity and are therefore set by the mass scale of the progenitor \citep[e.g.,][]{johnston98, apw14}. If $(\bs{x}_0,\bs{v}_0)$ are the parent orbit initial conditions at pericenter, then $(\delta\bs{x}_i,\delta\bs{v}_i)$ is the deviation vector of the $i$th particle and the magnitude of the offsets are assumed to be Normally distributed away from the parent orbit:
\begin{align}
	\delta\bs{x}_i &\sim \mathcal{N}(0, r_{\rm tide})\\
	\delta\bs{v}_i &\sim \mathcal{N}(0, \sigma_v)\\
	r_{\rm tide} &= \|\bs{x}_0\| \left(\frac{m}{3M_{\rm enc}(\|\bs{x}_0\|)}\right)^{1/3} \\
	\sigma_v &= \|\bs{v}_0\|\left(\frac{m}{3M_{\rm enc}(\|\bs{x}_0\|)}\right)^{1/3} 
\end{align}
where $M_{\rm enc}(r)$ is the mass enclosed of the host potential within radius $r$, $m$ is the mass scale of the debris, and $\|\cdot \|$ is the Euclidean norm. For this work, we take $m=10^4~\msun$ to represent globular-cluster-like progenitors, and use the spherically-averaged enclosed mass to estimate the above debris scales.

As the particle ensembles evolve in time, we follow a method similar to that of \citet[][hereafter MV96]{merritt96} to quantify the degree of mixing of the ensemble by computing the ``distance'' between the configuration space density of the ensemble at a given time, $\rho(\bs{x},t)$, with that of a distribution that is uniformly distributed over the energy hypersurface, $\rho_{E_0}(\bs{x})$, where $E_0$ is the energy of the parent orbit and thus hypersurface of interest. Our method differs in a few ways from that of MV96. First, MV96 use isoenergy ensembles of particles for following the density because this is the only way to guarantee that, if chaotic, nearly all orbits in the ensemble will lie on chaotic orbits. We instead use ensembles with small but finite spreads in energy; we still expect that most ensemble particle orbits around a parent chaotic orbit will reach a fully mixed state but now the final distribution will have some thickness around this surface. Around a typical regular orbit, each individual orbit will fill its torus ergodically --- a surface with lower dimension than the energy hypersurface --- and thus, a small enough ensemble should never reach the fully mixed state as defined above because it will remain confined to a subspace of lower dimensionality. In practice, only strongly chaotic systems will get close to $\rho_{E_0}(\bs{x})$ in a reasonable amount of time; we expect the rate to depend on how stochastic the parent orbit is (e.g., as measured by the Lyapunov time or frequency diffusion rate) and also on how ``trapped'' the chaotic region is (a small chaotic region wedged between stable tori will take longer to appear chaotic, even if internally the frequency diffusion rate is large). However, the difference in configuration space density between an ensemble undergoing chaotic mixing versus an ensemble around a regular orbit may be detectable over a much shorter timescale. The fully-mixed, normalized, configuration-space density that fills a given energy hypersurface, $E_0$, is easily found by projecting the simple distribution function, $f$, into configuration space:
\begin{align}
	f(\bs{x}, \bs{v}) &= f_0 \, \delta(H(\bs{x}, \bs{v}) - E_0)\\
	\rho_{E_0}(\bs{x}) &= \int \dd^3v \, f(\bs{x}, \bs{v})\\
	&= \int \dd^3v \, f_0 \, \delta(H - E_0)\\
	&= 4\pi \int \dd v \, v^2 \, f_0 \, \delta\left(v^2/2 + \Phi(\bs{x}) - E_0\right)\\
	&= A \sqrt{E_0 - \Phi(\bs{x})}\label{eq:mixeddensity}
\end{align}
where $A$ is a constant determined such that $\int \dd^3\bs{x} \, \rho_{E_0}(\bs{x}) = 1$, and $\delta(\cdot)$ is the Dirac delta function.

@@

\begin{align}
	E[\rho(\bs{x},t_j)] &= \int \rho(\bs{x},t_j) \, \rho(\bs{x},t_j) \, \dd \bs{x}\\
	&\approx \frac{1}{N}\sum \rho(\bs{x}_i,t_j)
\end{align}
So we can actually estimate the mean density by evaluating at the particles and taking the mean.
